---
title: "Antibiotics"
author: "Kelly Brennan"
date: 2020-07-28
output:
  github_document:
    toc: true
---

*Purpose*: To create an effective visualization, we need to keep our *purpose* firmly in mind. There are many different ways to visualize data, and the only way we can judge efficacy is with respect to our purpose.

In this challenge you'll visualize the same data in two different ways, aimed at two different purposes.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)
library(ggrepel)

```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization
<!-- -------------------------------------------------- -->

## Purpose: Compare Effectiveness
<!-- ------------------------- -->

__q1__ Create a visualization of `df_antibiotics` that helps you to compare the effectiveness of the three antibiotics across all the bacteria reported. Can you make any broad statements about antibiotic effectiveness?

```{r q1-task}
## TASK: Create your visualization

# (1) Organizing the data, so that the anitibotic is a variable in the dataset
df_mic <- 
  df_antibiotics %>% 
    pivot_longer(
      names_to = "antibiotic",
      values_to = "mic",    cols = penicillin:neomycin, 
    )
# print(df_longer)
df_human <- df_mic %>% filter(mic <= 0.01)

# (2) Plot
ggplot(data = df_mic, aes(x = bacteria, y = mic)) +
  geom_point(aes(color = antibiotic)) + 
  geom_point(data = df_human, shape = 3) +
  coord_flip() +
  scale_y_log10()+
  geom_vline(xintercept = 0.01, color = "black") #Why isn't this working?
  
```

**Observations**:

- Record your observations here
- Can you make any broad statements about antibiotic effectiveness?
- It's rather surprising that the three anitbiotics are not more effective against the bacteria. In creating this graph, I anticipated that most of them would be effective, but am proven wrong.
- Of the three antibiotics, penicillin is effective against the most different types of bacteria. Perhaps that is not surprising because it's the only anitbiotic that I've heard of before.
- I am surprised that streptomycin is only effective against Baccilus anthracis. I wonder why it is "useful" if all of the other antibiotics are effecvtive for the bacteria too. In a clinical setting, maybe it would be 'more' effective to use penicillin (but a lot of people are alergic).

## Purpose: Categorize Bacteria
<!-- ------------------------- -->

The *genus* of a living organism is a human categorization, based on various characteristics of the organism. Since these categories are based on numerous factors, we will tend to see clusters if we visualize data according to relevant variables. We can use these visuals to categorize observations, and to question whether given categories are reasonable!

__q2__ Create a visualization of `df_antibiotics` that helps you to categorize bacteria according to the variables in the data. Document your observations on how how clusters of bacteria in the variables do---or don't---align with their *genus* classification.

```{r q2-task}
## TASK: Create your visualization
# (1) Organizing the data more and making it tidy
df_genus <- 
  df_mic %>% 
    separate(
      col = bacteria, 
      into = c("genus", "species"), 
      sep = " ",
    )
print(df_genus)

#Filter for effectiveness
df_pass <- df_genus %>% filter(mic <= 0.01)


#(2) making a visualization
ggplot(df_genus, aes(fct_reorder(species, genus), y = mic)) + 
  geom_point(aes(color = genus)) + 
  geom_point(data = df_pass, aes(color = genus), shape = 3, size = 3) +
  coord_flip() + 
  scale_y_log10() +
  facet_wrap(~ antibiotic)


#filter for the types of genus with more than one bacteria species
df_genus %>% 
  filter(genus %in% c("Streptococcus", "Staphylococcus", "Salmonella")) %>% 
  ggplot(aes(x = mic, y = fct_reorder(species, genus))) + 
  geom_point(aes(color = antibiotic)) + 
  facet_wrap(~ genus)


```

**Observations**:

- Pretty colors! 
- There are only three genus with more than on species in this dataplot: Streptococcus, Staphyloccus, and Salmonella.
- When remove the log scale axis, I can see that none of the outliers are from the same genus.
- It does appear that bacteria from the same genus have mic values that are not super far away from each other. I decided to make the second plot to isolate bacteia from the same genus, and confirm that they have roughly simlar effectiveness with the same antibiotic.


# References
<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." /American Scientist/ (2009) [link](https://www.americanscientist.org/article/thats-funny)
